Synthesis of Linear Functional Programs

### 2021-05-09 article/article.tex
touch ups

\section{Background}

@@

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
approach very similar to our own.  The work employs so-called graded
modal types, which is an refinement of pure linear types -- a more
\emph{fine-grained} version, since it allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions.  Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
resource management approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09} -- the model we'll be using.

They also use focusing as a solution to trim down search space and
to ensure that synthesis only produces well-typed programs, and
our solution will use the same literature as a base.  However, since
their underlying logic is \emph{modal} rather than purely
\emph{linear}, it lacks a clear correspondence with concurrent
session-typed programs~\cite{}, which is a crucial avenue of future
work. Moreover, their use grading effectively requires constraint solving to
be integrated with the synthesis procedure, which can limit the
effectiveness of the overall approach as one scales to more
sophisticated settings (e.g.~refinement types).

@@

We start with the definition of our type system and simply-typed functional language
with linear types ($\lolli$, $\tensor$, $1$, $\with$, $\oplus$, $\bang$, $Bool?$).
Through the Curry-Howard correspondence, we'll formulate the synthesis of a
linear type in our functional lanugage as proof search in linear logic.
Since \emph{natural deduction} is ``natural'' in practice, and thus, harder to streamline and reason about,
we'll use the equivalent \emph{sequent calculus} to formulate our inference rules.
In sequent calculus, all our rules can be read in a \emph{bottom-up} manner, simplifying the reasoning for the synthetizer.
Therafter, a way to reason about our resource contexts is needed, and the Hodas and Miller's? input-output model~\cite{} comes into play.
According to it, instead of a single context that we must split into parts when proving the judgments
above the inference rule line, we'll utilize an \emph{input} and an \emph{output} context -- to prove $A\tensor B$ with $\Delta$, prove $A$ with $\Delta$, consuming the needed resources, and output a $\Delta'$, prove $B$ with $\Delta'$ and the output the remaining resources in $\Delta''$, and finally, after having proven $A\tensor B$, output $\Delta''$, which might be used by another judgment.
With the input/output contexts, and the sequent calculus formulation of the inference rules, we're on a good path to synthetize programs.
However, even though proof-search here might seem direct, it still encompasses a lot of non-determinism, e.g., which rule to apply at a certain point?
To this we add focusing~\cite{} -- a method created to trim down the search space of valid proofs in a linear logic, by conducting the
proof-search first with invertible rules, and then by focusing where no more direct (invertible) rules can be applied.
With focusing we tame some of non-determinism from the synthesis aspect. \todo{talvez algo mais sobre focusing aqui.}
We might, later on, add refinement types also as a way to constrain the valid programs space.
Additionally, we might need to discuss strategies to synthetize recursive functions as well.

>>>>

\section{Background}\label{sec:background}

@@

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
approach very similar to our own.  The work employs so-called graded
modal types, which is an refinement of pure linear types -- a more
\emph{fine-grained} version, since it allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions.  Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
resource management
approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
-- the model we will be using.

They also use focusing as a solution to trim down search space and to
ensure that synthesis only produces well-typed programs, and our
solution will use the same literature as a base.  However, since their
underlying logic is \emph{modal} rather than purely \emph{linear}, it
lacks a clear correspondence with concurrent session-typed
programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
which is a crucial avenue of future work. Moreover, their use grading
effectively requires constraint solving to be integrated with the
synthesis procedure, which can limit the effectiveness of the overall
approach as one scales to more sophisticated settings (e.g.~refinement
types).

@@

We start with the definition of our type system and simply-typed
functional language with linear function, sum and product types
($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
the Curry-Howard correspondence, we'll formulate the synthesis of
typed programs in our functional language as bottom-up proof-search in
linear logic.

Since proof rules in \emph{natural deduction} do not entail a
proof-search strategy, we will use the equivalent \emph{sequent
  calculus} system to formulate our typing rules in a way that is
amenable to proof search.
%
In sequent calculus, rules can be naturally understood in a
\emph{bottom-up} manner, simplifying the reasoning for the
synthesizer. Moreover, the linear nature of our system requires a way
to algorithmically reason about resource contexts: for instance, if we
read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
must non-deterministically guess how to correctly split ambient
resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
$B$. To solve this issue we will adopt the resource-management
approach
of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
Instead of requiring non-deterministic guesses of resource splits
during proof search, we will make use of an \emph{input} and an
\emph{output} context, leading to the following general strategy: to
prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
consuming some subset of $\Delta$, and produce as output the leftover
resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
remaining resources $\Delta''$; finally, after having proven
$A\tensor B$, output $\Delta''$, for subsequent derivations.

With the use of input/output contexts, and the sequent calculus
formulation of the inference rules, we draw closer to an effective
strategy for bottom-up proof search and so of program synthesis.
%

However, the search process still encompasses significant
non-determinism, e.g., at any given point, many proof rules are
applicable in general. To address this we draw on the technique of
focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} --
a method created to trim down the search space of valid proofs in
linear logic, by eagerly applying invertible rules (i.e.~rules whose
conclusion implies the premises), and then by ``focusing'' on a single
connective when no more direct (invertible) rules can be applied. If
the search is not successful, the procedure backtracks and another
connective is chosen as the focus.
%
Focusing eliminates all of the
``don't care'' non-determinism from proof search, since the order in
which invertible rules are applied does not affect the outcome of the
search, leaving only the non-determinism that pertains to unknowns (or
``don't know'' non-determinism), identifying precisely the points at
which backtracking search is necessary.

With the basic framework in place, we will consider additional
(simply-typed) language constructs and synthesis of program fragments
or \emph{partial synthesis} -- given a program with a typed hole,
synthesize a term to fill out the hole accordingly. We will
subsequently also consider synthesis of recursive functions and,
might, later on, add refinement types also as a way to constrain the
valid programs space.
